# Google Data Analytics (Cyclistic) Capstone Project

---
title: "capstone_case-study_1"
author: "Abhirup"
date: "2022-08-27"
output: word_document
---

#Introduction

This is my version of the Google Data Analytics Capstone - Case Study 1. The full document to the case study can be found in the Google Data Analytics Capstone: Complete a Case Study course.

For this project this steps will be followed to ensure its completion:

*It will follow the steps of the data analysis process: Ask, prepare, process, analyze, share, and act.
*Each step will follow its own roadmap with:
**Code, if needed on the step.
**Guiding questions, with answers.
**Key tasks, as a checklist.
**Deliverable, as a checklist.

#Process

This step will prepare the data for analysis. All the csv files will be merged into one file to improve workflow

#Code

##Dependences
The main dependencie for the project will be tidyverse.

```{r}
# Install packages
# if (!require(package)) install.packages('package')
```

#Load Libraries

```{r setup, include=FALSE}
#load libraries 
library(tidyverse) #calculations
library(lubridate) #dates 
library(hms)#time
library(dplyr) #cleaning
library(skimr) #summarize
library(data.table) #exporting data frame
```

#Data

The data is on an AWS server where it is easily downloadable and named correctly. I downloaded the previous 12 months data and stored it locally for the next steps in the analysis processes. It is organized by year and Fiscal Quarters. The data is reliable and original since it comes from the company.It is comprehensive, current, and cited. The data source is the company so everything about the users personal information is hidden or kept private to the company only.

Note: that data-privacy issues prohibit you from using riders’ personally identifiable information. This means that you won’t be able to connect pass purchases to credit card numbers to determine if casual riders live in the Cyclistic service area or if they have purchased multiple single passes.

###Data Set URL: https://divvy-tripdata.s3.amazonaws.com/index.html


#STEP 1: COLLECT DATA

```{r Laod Data, results="hide"}
#load original .csv files, a years worth of data from August 2020 to July 2021
Q1_2020_df <- read_csv("cyclistic_data/Divvy_Trips_2020_Q1.csv")
apr_2020_df <- read_csv("cyclistic_data/202004-divvy-tripdata.csv")
may_2020_df <- read_csv("cyclistic_data/202005-divvy-tripdata.csv")
jun_2020_df <- read_csv("cyclistic_data/202006-divvy-tripdata.csv")
jul_2020_df <- read_csv("cyclistic_data/202007-divvy-tripdata.csv")
aug_2020_df <- read_csv("cyclistic_data/202008-divvy-tripdata.csv")
sep_2020_df <- read_csv("cyclistic_data/202009-divvy-tripdata.csv")
oct_2020_df <- read_csv("cyclistic_data/202010-divvy-tripdata.csv")
nov_2020_df <- read_csv("cyclistic_data/202011-divvy-tripdata.csv")
dec_2020_df <- read_csv("cyclistic_data/202012-divvy-tripdata.csv")

jan_2021_df <- read_csv("cyclistic_data/202101-divvy-tripdata.csv")
feb_2021_df <- read_csv("cyclistic_data/202102-divvy-tripdata.csv")
mar_2021_df <- read_csv("cyclistic_data/202103-divvy-tripdata.csv")
apr_2021_df <- read_csv("cyclistic_data/202104-divvy-tripdata.csv")
may_2021_df <- read_csv("cyclistic_data/202105-divvy-tripdata.csv")
jun_2021_df <- read_csv("cyclistic_data/202106-divvy-tripdata.csv")
jul_2021_df <- read_csv("cyclistic_data/202107-divvy-tripdata.csv")
aug_2021_df <- read_csv("cyclistic_data/202108-divvy-tripdata.csv")
sep_2021_df <- read_csv("cyclistic_data/202109-divvy-tripdata.csv")
oct_2021_df <- read_csv("cyclistic_data/202110-divvy-tripdata.csv")
nov_2021_df <- read_csv("cyclistic_data/202111-divvy-tripdata.csv")
dec_2021_df <- read_csv("cyclistic_data/202112-divvy-tripdata.csv")

jan_2022_df <- read_csv("cyclistic_data/202201-divvy-tripdata.csv")
feb_2022_df <- read_csv("cyclistic_data/202202-divvy-tripdata.csv")
mar_2022_df <- read_csv("cyclistic_data/202203-divvy-tripdata.csv")
apr_2022_df <- read_csv("cyclistic_data/202204-divvy-tripdata.csv")
may_2022_df <- read_csv("cyclistic_data/202205-divvy-tripdata.csv")
jun_2022_df <- read_csv("cyclistic_data/202206-divvy-tripdata.csv")
jul_2022_df <- read_csv("cyclistic_data/202207-divvy-tripdata.csv")

```

```{r Fix Data Type, results="hide"}
#fixing datatypes
# This step was necessary because the below imported CSV data was not in correct date-time format.
# Wrong format could through error in later part.

Q1_2020_df <- Q1_2020_df %>% mutate(started_at = dmy_hm(started_at))
Q1_2020_df <- Q1_2020_df %>% mutate(ended_at = dmy_hm(ended_at))

apr_2020_df <- apr_2020_df %>% mutate(started_at = dmy_hm(started_at))
apr_2020_df <- apr_2020_df %>% mutate(ended_at = dmy_hm(ended_at))
```


#WRANGLE DATA AND COMBINE INTO A SINGLE FILE

```{r, results="hide"}
#merge all of the data frames into one year view
all_trips <- rbind(Q1_2020_df, 
                   apr_2020_df,
                   may_2020_df,
                   jun_2020_df,
                   jul_2020_df,
                   aug_2020_df,
                   sep_2020_df,
                   oct_2020_df,
                   nov_2020_df,
                   dec_2020_df,
                   
                   jan_2021_df,
                   feb_2021_df,
                   mar_2021_df,
                   apr_2021_df,
                   may_2021_df,
                   jun_2021_df,
                   jul_2021_df,
                   aug_2021_df,
                   sep_2021_df,
                   oct_2021_df,
                   nov_2021_df,
                   dec_2021_df,
                   
                   jan_2022_df,
                   feb_2022_df,
                   mar_2022_df,
                   apr_2022_df,
                   may_2022_df,
                   jun_2022_df,
                   jul_2022_df)
```

#Remove temorary data frames

```{r, results="hide"}
#remove individual month data frames to clear up space in the environment 
remove(Q1_2020_df,
       apr_2020_df,
       may_2020_df,
       jun_2020_df, 
       jul_2020_df, 
       aug_2020_df, 
       sep_2020_df, 
       oct_2020_df, 
       nov_2020_df,
       dec_2020_df, 
       
       jan_2021_df,
       feb_2021_df,
       mar_2021_df,
       apr_2021_df,
       may_2021_df,
       jun_2021_df,
       jul_2021_df,
       aug_2021_df,
       sep_2021_df,
       oct_2021_df,
       nov_2021_df,
       dec_2021_df,
       
       jan_2022_df,
       feb_2022_df,
       mar_2022_df,
       apr_2022_df,
       may_2022_df,
       jun_2022_df,
       jul_2022_df)
```


#Summerize Data 

```{r Summerize Data}
#Summerize Data
colnames(all_trips)  #List of column names

```

```{r}
nrow(all_trips)  #How many rows are in data frame?

```

```{r}
dim(all_trips)  #Dimensions of the data frame?

```

```{r}
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)

```

```{r}
str(all_trips)  #See list of columns and data types (numeric, character, etc)

```

```{r}
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```


```{r}
glimpse(all_trips) #summary
```

#Create dataframe for Date-Time Analysis

```{r}
#create new data frame to contain new columns
cyclistic_datetime <- all_trips
```

```{r}
#calculate ride length by subtracting ended_at time from started_at time and converted it to minutes
cyclistic_datetime$ride_length <- difftime(all_trips$ended_at, all_trips$started_at, units = "mins")
```

```{r}
#create columns: day of week, month, day, year, time, hour
cyclistic_datetime$date <- as.Date(cyclistic_datetime$started_at) #default format is yyyy-mm-dd, use start date
cyclistic_datetime$day_of_week <- wday(all_trips$started_at) #calculate the day of the week
cyclistic_datetime$day_of_week <- format(as.Date(cyclistic_datetime$date), "%a")
cyclistic_datetime$month <- format(as.Date(cyclistic_datetime$date), "%m")#create column for month
cyclistic_datetime$day <- format(as.Date(cyclistic_datetime$date), "%d") #create column for day
cyclistic_datetime$year <- format(as.Date(cyclistic_datetime$date), "%Y") #create column for year
cyclistic_datetime$time <- format(as.Date(cyclistic_datetime$date), "%H:%M:%S") #format time as HH:MM:SS
cyclistic_datetime$time <- as_hms((cyclistic_datetime$started_at)) #create new column for time
cyclistic_datetime$hour <- hour(cyclistic_datetime$time) #create new column for hour
```

```{r, results="hide"}
#create column for different seasons: Spring, Summer, Fall, Winter
cyclistic_datetime <-cyclistic_datetime %>% mutate(season = 
                                             case_when(month == "03" ~ "Spring",
                                                       month == "04" ~ "Spring",
                                                       month == "05" ~ "Spring",
                                                       month == "06"  ~ "Summer",
                                                       month == "07"  ~ "Summer",
                                                       month == "08"  ~ "Summer",
                                                       month == "09" ~ "Fall",
                                                       month == "10" ~ "Fall",
                                                       month == "11" ~ "Fall",
                                                       month == "12" ~ "Winter",
                                                       month == "01" ~ "Winter",
                                                       month == "02" ~ "Winter")
                                                    )
```

```{r, results="hide"}
#create column for different time_of_day: Night, Morning, Afternoon, Evening
cyclistic_datetime <-cyclistic_datetime %>% mutate(time_of_day = 
                                             case_when(hour == "0" ~ "Night",
                                                       hour == "1" ~ "Night",
                                                       hour == "2" ~ "Night",
                                                       hour == "3" ~ "Night",
                                                       hour == "4" ~ "Night",
                                                       hour == "5" ~ "Night",
                                                       hour == "6" ~ "Morning",
                                                       hour == "7" ~ "Morning",
                                                       hour == "8" ~ "Morning",
                                                       hour == "9" ~ "Morning",
                                                       hour == "10" ~ "Morning",
                                                       hour == "11" ~ "Morning",
                                                       hour == "12" ~ "Afternoon",
                                                       hour == "13" ~ "Afternoon",
                                                       hour == "14" ~ "Afternoon",
                                                       hour == "15" ~ "Afternoon",
                                                       hour == "16" ~ "Afternoon",
                                                       hour == "17" ~ "Afternoon",
                                                       hour == "18" ~ "Evening",
                                                       hour == "19" ~ "Evening",
                                                       hour == "20" ~ "Evening",
                                                       hour == "21" ~ "Evening",
                                                       hour == "22" ~ "Evening",
                                                       hour == "23" ~ "Evening")
                                                      )
```

```{r, results="hide"}
#create a column for the month using the full month name
cyclistic_datetime <-cyclistic_datetime %>% mutate(month = 
                                             case_when(month == "01" ~ "January",
                                                       month == "02" ~ "February",
                                                       month == "03" ~ "March",
                                                       month == "04" ~ "April",
                                                       month == "05" ~ "May",
                                                       month == "06" ~ "June",
                                                       month == "07" ~ "July",
                                                       month == "08" ~ "August",
                                                       month == "09" ~ "September",
                                                       month == "10" ~ "October",
                                                       month == "11" ~ "November",
                                                       month == "12" ~ "December"
                                                       )
                                            )
```

#View Uncleaned Data

```{r}
#----------View UnCleaned Data
head(cyclistic_datetime)

```


#Data cleaning

```{r}
#clean the data
cyclistic_datetime <- cyclistic_datetime %>% rename(bike_type = rideable_type) #Renaming Columns for more understandability 
cyclistic_datetime <- na.omit(cyclistic_datetime) #remove rows with NA values
cyclistic_datetime <- distinct(cyclistic_datetime) #remove duplicate rows 
cyclistic_datetime <- cyclistic_datetime[!(cyclistic_datetime$ride_length <=0),] #remove where ride_length is 0 or negative
cyclistic_datetime <- cyclistic_datetime %>%  #remove columns not needed: ride_id, start_station_id, end_station_id, start_lat, start_long, end_lat, end_lng
  select(-c(ride_id, start_station_id, end_station_id))
#Data cleaned Up
message("Cleaned Up ", nrow(all_trips)-nrow(cyclistic_datetime), " Rows")
```

#Final Data

```{r}
#view the final data
str(cyclistic_datetime)
```

```{r}
glimpse(cyclistic_datetime)
```

#ANALYSIS

```{r}
#-----------------------------------------TOTAL RIDES--------------------------------------

#total number of rides
nrow(cyclistic_datetime)
```

#Data distribution
Here we want to try to answer the most basic questions about how the data is distributed.


##Casuals vs members
How much of the data is about members and how much is about casuals?

```{r}
#-----------------MEMBER TYPE---------------------
cyclistic_datetime %>%
  group_by(member_casual) %>% 
  count(member_casual)

```
```{r} 
#-----------------Plot MEMBER TYPE---------------------
ggplot(cyclistic_datetime, aes(member_casual, fill=member_casual)) +
  geom_bar() +
  labs(x="Casuals x Members", title="Chart 01 - Casuals x Members distribution")
```

```{r}
#----------------TYPE OF BIKE---------------------

#total rides by member type 
cyclistic_datetime %>%
  group_by(member_casual, bike_type) %>% 
  count(bike_type)

```

#Grouped barchart
###Ref: https://r-graph-gallery.com/48-grouped-barplot-with-ggplot2.html

Plotting Bar chart by Grouping bike_types w.r.t Member_casual

```{r}
#Plot
cyclistic_datetime %>% 
  group_by(member_casual, bike_type) %>% 
  summarise(number_of_rides = n()) %>% 
  ggplot(aes(x = bike_type, y = number_of_rides, fill = member_casual)) +
  geom_bar(position = "dodge", stat='identity')
```

#Plotting - Distribution by time of the day


```{r}
cyclistic_datetime %>%
    ggplot(aes(time, fill=member_casual)) +
    labs(x="Time of the day", title="Distribution by hour of the day") +
    geom_bar()  
```


```{r}
cyclistic_datetime %>%
    ggplot(aes(hour, fill=member_casual)) +
    labs(x="Time of the day", title="Distribution by hour of the day") +
    geom_bar()
```

```{r}
cyclistic_datetime %>%
    ggplot(aes(hour, fill=member_casual)) +
    geom_bar() +
    labs(x="Hour of the day", title="Chart 05 - Distribution by hour of the day divided by weekday") +
    facet_wrap(~ day_of_week)
```


```{r}
#----------------------TIME OF DAY-----------------------

#total rides by member type 
cyclistic_datetime %>%
  group_by(member_casual) %>% 
  count(time_of_day)
```
```{r}
cyclistic_datetime %>%
    ggplot(aes(time_of_day, fill=member_casual)) +
    labs(x="Time of the day", title="Distribution by Time of the day") +
    geom_bar()
```


```{r}
#----------------DAY OF THE WEEK------------------

#total rides by member type
cyclistic_datetime %>%
  group_by(member_casual) %>% 
  count(day_of_week)

```

```{r}
#total rides 
cyclistic_datetime %>%
  count(day_of_week)
```



```{r}
cyclistic_datetime %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(day_of_week)%>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```


```{r}
#---------------------MONTH-----------------------

#total rides by member type 
cyclistic_datetime %>%
  group_by(member_casual) %>% 
  count(month) %>% 
  print(n = 24) #lets you view the entire tibble

#total rides
cyclistic_datetime %>%
  count(month)
```

#Plotting Season Data

```{r}
#--------------------SEASON-----------------------

#-----all seasons-------

#total rides by member type
cyclistic_datetime %>%
  group_by(season, member_casual) %>% 
  count(season)
```

```{r}
#total rides
cyclistic_datetime %>%
  group_by(season) %>% 
  count(season)
```

```{r}
cyclistic_datetime %>%
    ggplot(aes(month, fill=member_casual, color=season)) +
    labs(x="Season", title="Distribution by Season") +
    geom_bar()
```


